Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2022-02-22 Paper/Paper.tex

\usepackage[backend=biber,style=authoryear,
sorting=ynt,citestyle=authoryear]{biblatex}
\addbibresource{papercitations.bib}
\usepackage{setspace}
\doublespacing
\addtolength{\skip\footins}{2pc plus 5pt}

@@

The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making. Their interaction has direct impacts across industries; job displacement, creation, and satisfaction are all prevalent reality for many workers (\textcolor{red}{PUT SOME CITATIONS HERE}). Additionally, in some cases, labor market changes that take place for one job can have rippling effects to populations of people. One such job is that of a physician, where physician decisions have direct and indirect effects on patients receiving care. For example, a physician experiencing burnout from work pressures may treat patients differently and lead to patients feeling uncomfortable seeking care (\textcolor{red}{FIND A CITE FOR THIS)}, or increased administrative burdens could lead to less time with patients and exacerbated access to care issues for those in rural settings (\textcolor{red}{FIND CITE FOR THIS}). There may also be positive effects on patients from physician labor market changes, such as increased access to care that comes from more physicians entering the market, or less harmful mistakes that arise from standardizing care given. For these reasons, I investigate physician labor markets under a major technology change in the U.S. healthcare system, the implementation of electronic health care record systems, which have been emphasized as a cause of burnout and may induce labor market changes in the occupation. \textcolor{red}{Rethink how I frame the examples, but MAKE SURE I HAVE CITATIONS FOR THESE THINGS}

Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggesions for care. EHRs have have become increasingly relevant in the U.S. since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The movement towards digitization in health care was expected to have immediate and substantial impacts on efficiency of care. A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}). Such expectations led the U.S. government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}. The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), evidence of the nationwide movement towards digitization.

Physicians, as primary users of this technology, have a meaningful role in whether potential benefits of EHRs are realized. However, with the rapid implementation of EHRs, the day to day life of physicians changed drastically. In interviews, groups of physicians reported that when using EHRs they are less satisfied with their job and have higher burnout and higher stress levels. Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices. Further, the short and long run impacts of EHRs on physician productivity could vary greatly.  I use modern event study approaches, including continuous treatment, to understand the impact of EHR implementation in hospitals on physician labor market outcomes: retirement, where to work, and productivity in hospitals.

Using CMS shared patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct physician-level data on hospitalists and other general practice physicians connected to hospitals from 2009-2017. The data captures the extent to which the physician in exposed to EHRs in hospitals and various labor market outcomes, along with demographic information such as zip code, age, and gender. With this data I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), and (3) the number of patients seen by the physician. Common to each specification is the use of treatment variables capturing a physicians' level of EHR exposure in hospitals.  For retirement, I focus primarily on the sample of older physicians who are closer to the margin of retirement, since the opportunity cost of retiring for a young physician is very high and the cost of using an EHR is unlikely to exceed it. Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that \textcolor{red}{put exact findings here (if no effect, take out the first part of this sentence)}.

Given that there is a population of both younger and older physicians who stay in the labor force (\textcolor{red}{it may be that all of them stay in the market}), the next outcome to consider is whether physicians switch work settings as a result of EHR implementation in hospitals. That is, do physicians switch from primarily hospital based work to smaller office based settings, or do they switch hospital locations where the technology looks different?  Here, I limit the sample to any physicians who stay in the labor force after EHR exposure. I find that \textcolor{red}{put exact findings here and their implications (are physicians trying to shift away from an autonomous standard of care they have to abide by?)}.

These findings imply that there are still physicians who remain employed by the same hospital after EHR implementation. Thus, I study the effect of the technology on productivity within hospitals, where productivity is measured by the number of patients seen. The population of physicians left in this analysis are those who do not retire or switch to an office based setting. I find that productivity does not change in the immediate years after implementation, and that it takes two years after exposure to see productivity gains. After 5 years of exposure, a physician is seeing on average 11\% more patients relative to before EHRs. This result speaks to the disconnect between expected effects and realized effects of EHRs which is discussed in the health information technology literature. Since more patients are being seen by physicians, efficiency and access to care have both increased while costs (Medicaid patients) and quality remain unchanged. \textcolor{red}{This paragraph needs some work as my results have changed since getting the MDPPAS data and I need to push on other mechanisms besides productivity - if other physicians left and work had to be picked up or billing activity increased or demand increased (maybe due to ACA?)}

@@

The effects of EHR implementation on costs and quality of healthcare prior to 2010 have been studied extensively. Despite many case studies and hospital-level analysis that indicate s large decrease in mortality rates (\cite{Buntin2011TheResults}), more recent economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}). Further, there have been no significant cost decreases due to EHRs in the short or long run. If any cost reductions are realized at all, they are at least 6 years after implementation (\cite{dranove2014trillion}). As discussed earlier, these findings are puzzling due to the expected benefits of EHRs. This paper contributes to our knowledge of this area in re-framing past finding into a patient access to care problem instead of an efficiency problem.

In 2009, the

@@

EHRs have been an important feature of healthcare since the 1980s and 1990s. Early on, this technology was mainly produced and used by academic medical centers whose primary goal was to create efficiency in billing and scheduling. Before the rapid development of computers, physicians were not even able to interface directly with EHRs, and thus were not drastically affected. The cost of utilizing an EHR was extremely large and increasing in the amount of capabilities, which led to the widespread view that they were only complements to paper records and not feasible as a substitute. The feasibility of using only EHRs increased as computers became more portable, creating what is known as the "physician workspace", a computer station for the physician to interface directly with EHRs. However, physicians kept the view that the technology was purely complementary to paper due to burdensome data entry. Automation in data entry was non-existent, making it extremely time consuming for the user. Even when automation was developed for certain data, policy required users to double check all of the data in the EHR for errors.

The HITECH Act was passed in 2009, setting aside government funds to subsidies hospitals who used EHRs according to certain guidelines. These requirements include having at least 80\% of patients in the system, asking certain questions to patients, and protecting the security of the system. Immediately, EHR companies took advantage of the legislation and began marketing campaigns centered around their system being eligible to earn the hospital subsidies. Hospitals began receiving payments for the first stage of meaningful use in 2011. This led to a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}). \textcolor{red}{maybe put HealthIT US map that shows hospitals who received incentive payment?} While EHRs are widely implemented in hospitals by 2015, some physicians still continue to report frustration. "Some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}). This shows the continued hesitancy to utilize EHRs that physicians exhibit.

@@

Using various linked datasets, I construct a physician-level panel spanning from 2009-2015 which measures physicians' exposure to EHRs over time and other relevant characteristics. The different data sets used to construct this panel are described below.

The main data I utilize is the CMS Shared Patients Data. For a pair of National Provider Identifiers (NPIs), it records the number of patients who bill both of the entities under Medicaid in the same day. For example, if a Medicaid patient is referred to a specialist by a primary care physician, then those two physicians have a shared patient in common. I limit the entities to only include the number of shared patients for physician-hospital pairs. I care specifically about physicians who have a close working relationship with hospitals, preferably those who do rounds within at least one hospital. To achieve this, I limit the qualifications for a sufficient physician-hospital pair. First, I only consider doctors who have a primary care role, excluding any specialists. The reason for this is that a primary care physician who does rounds in a hospital is likely to be primarily employed by that hospital (not brought in from a practice). Another reason for this is that primary care doctors interact directly with EHRs during checkups, unlike a surgeon. Second, I limit pairs to those which have a substantial number of patients billed together in the same day. This is to avoid including the potential physician who works in an office but happens to send a small number of patients to a nearby hospital. I consider a substantial number of patients to be at least 30 patients bill both the physician and the hospital in the same day in at least one year, \textcolor{red}{where the exact number chosen is explored in the appendix}. The physicians remaining in the data are likely doing a number of rounds physically inside hospitals, since it is unlikely that many patients will see a primary care doctor in a private office and then immediately go to a hospital in the same day.

@@

@@

\vspace{5mm}
\begin{figure}[ht]
\caption{Frequency of Data Assistant Enumeration by Year}
    \includegraphics[scale=.5]{Objects/dataassistant_histogram.pdf}
    \label{fig:dataassistant_histogram}
\end{figure}

>>>>

\usepackage[backend=biber,style=authoryear,
sorting=ynt,citestyle=authoryear]{biblatex}
\addbibresource{papercitations.bib}
\usepackage{setspace}
\singlespacing
\addtolength{\skip\footins}{2pc plus 5pt}

@@

The relationship between technological innovation and labor markets is a longstanding topic of interest in economics and in policy making. Their interaction has direct impacts across industries; job changes such as displacement, creation, and satisfaction are all prevalent for many workers (\cite{autor2003skill}, \cite{fallick1996review}, \cite{akerlof1988job}). Labor market changes for one job type may also have rippling effects on populations serviced by that industry. One such job is that of a physician, where physician choices and labor markets directly affect patients on margins of accessibility and quality of care received. For example, physician burnout is correlated with lower patient satisfaction (\cite{shanafelt2002burnout}), physician shortages are linked to higher mortality rates (\cite{gong2019higher}), and administrative regulations such as standardized checklists may be associated with lower mortality in surgical settings (\cite{treadwell2014surgical}). Thus, shocks to physician labor markets or work settings may have drastic implications for patients. I investigate physician labor markets under a major technology change in the U.S. healthcare system, the implementation of electronic health care record systems, which have been emphasized as a cause of burnout and may induce changes to physician labor market decisions.

Electronic Health Records (EHRs) are computerized medical records which include detailed accounts and notes of medical history, stored in advanced systems with additional capabilities such as providing suggesions for care. EHRs have have become increasingly relevant in the U.S. since 2008, in part after former President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The movement towards digitization in health care was expected to have immediate and substantial impacts on efficiency. A 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}). Such expectations led the U.S. government to incentivize the use of EHRs in hospitals with the the Health Information Technology for Economic and Clinical Health (HITECH) Act in 2008 (\cite{hitech})\footnote{This legislation subsidized hospitals which used EHRs “meaningfully”According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})}. The percentage of hospitals with the capability of using a basic EHR system went from 9 percent in 2008 to 84 percent in 2015 (\cite{stats}), evidence of the nationwide movement towards digitization.

Physicians, as primary users of this technology, have a meaningful role in whether potential benefits of EHRs are realized. However, with the rapid implementation of EHRs, the day to day life of physicians also changed rapidly. In interviews, physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular were found to “loathe the cumbersome, time-consuming data entry that comes with using EHRs.” (\cite{CollierBurnout}). The frustration of using a new technology raises the cost of working in specific hospitals, which may lead physicians on the margin to make changes in labor market choices such as exiting the labor force altogether. Further, the short and long run impacts of EHRs on physician productivity could vary greatly due to the immediate learning curve that subsides over time.  I use difference in difference with heterogeneous treatment timing to understand the effect of EHR implementation in hospitals on physician labor market outcomes: retirement, choice of work setting, and patients seen in the short vs. long run.

Using CMS Shared Patient data and Medicare Data on Provider Practice and Specialty (MD-PPAS), I construct a physician-level panel on hospitalists and other general practice physicians connected to hospitals from 2009-2017. The data captures the extent to which the physician in exposed to EHRs in hospitals, various labor market outcomes, and demographic information such as zip code, age, and gender. I analyze the impact of EHR exposure on the following physician decisions: (1) retirement, (2) where to physically work (changing hospitals or moving to office based settings), and (3) the number of patients seen by the physician. Common to each specification is the use of treatment variables capturing a physicians' level of EHR exposure in hospitals. I consider exposure on two dimensions: working in any hospital that implements an EHR, and working in a low-integrated hospital that implements an EHR.  To study retirement, I focus primarily on the sample of older physicians who are closer to the margin of retirement, since the opportunity cost of retiring for a young physician is very high and the cost of using an EHR is unlikely to exceed it. Contrary to previous work stating that physicians do not retire based on work environment (\cite{Bahrami2002}), I find that \textcolor{red}{put exact findings here. (The funny thing is, I find that young physicians "retire"... I need to figure out what this means and possibly rephrase the last couple sentences)}.

For those who stay in the labor force, there may be margin to change physicil place of work if physicians are unhappy with the specific loss of autonomy experienced at a given hospital. That is, do physicians switch from primarily hospital based work to smaller office based settings, or do they switch hospital locations where the technology looks different? To answer this question, I limit the sample to any physicians who stay in the labor force after EHR exposure and use the same event study framework to estimate the effect on whether the physician switches zip codes or increases office-based patients. I find that \textcolor{red}{put exact findings here and their implications (are physicians trying to shift away from an autonomous standard of care they have to abide by?)}.

Finally, I investigate the effect of the technology implementation on productivity within hospitals, measured by the number of patients seen. The population of physicians left in this analysis are those who do not retire or switch to an office based setting. I find that the number of patients seen does not change in the immediate years after implementation, and that it takes two years after exposure to see productivity gains. After 5 years of exposure, a physician is seeing on average 11\% more patients relative to before EHRs. This result speaks to the disconnect between expected effects and realized effects of EHRs which is discussed in the health information technology literature. Since more patients are being seen by physicians, efficiency and access to care have both increased while costs (Medicaid patients) and quality remain unchanged. \textcolor{red}{This paragraph needs some work as my results have changed since getting the MDPPAS data and I need to push on other mechanisms besides productivity - if other physicians left and work had to be picked up or billing activity increased or demand increased (maybe due to ACA?)}

@@

The effects of EHR implementation on costs and quality of healthcare prior to 2010 have been studied extensively. Despite many case studies and hospital-level analysis that indicate a large decrease in mortality rates (\cite{Buntin2011TheResults}), more recent economic research has showed that the technology has only improved health outcomes for patients with severe conditions, but have not led to improvements for the median patient (\cite{Agha2014TheCare}; \cite{McCullough2016HealthCoordination}; \cite{Meyerhoefer}). Further, there have been no significant cost decreases due to EHRs in the short or long run. If any cost reductions are realized at all, they are at least 6 years after implementation (\cite{dranove2014trillion}). As discussed earlier, these findings are puzzling due to the expected benefits of EHRs. This paper contributes to our knowledge of this area in re-framing past finding into a patient access to care problem instead of an efficiency problem.

\textcolor{red}{Expand both contribution and literature review and combine them to flow better.}

@@

EHRs have been an important feature of healthcare since the 1980s and 1990s. Early in its existence, the technology was mainly produced and used by academic medical centers whose primary goal was to create efficiency in billing and scheduling. Before the rapid development of computers, physicians were not even able to interface directly with EHRs, and thus were not drastically affected. The cost of utilizing an EHR was extremely large, and cost increased as the amount of capabilities increased, which led to the widespread view that EHRs were strictly complements to paper records and not a feasible substitute. The possibility of using solely EHRs increased as computers became more portable, creating what is known as the "physician workspace": a computer station for a physician to interface directly with EHRs to record patient updates. However, physicians kept the view that EHRs were purely complementary to paper due to burdensome data entry. Automation in data entry was non-existent, making it extremely time consuming for the user. Even when automation was developed for particular machines that performed monitoring, the hospital still held responsible for the accuracy of information and therefore required physicians to double check each data entry.

The HITECH Act was passed in 2009, setting aside government funds to subsidize hospitals who used EHRs according to certain guidelines. These requirements include having at least 80\% of patients in the system, recording answers to specific questions, and protecting the security of the system to ensure privacy. Immediately, EHR companies took advantage of the legislation and began marketing campaigns centered around their system being eligible to earn the hospital subsidies. Hospitals began receiving payments for the first stage of meaningful use in 2011. This led to a 75 percentage point increase in the number of hospitals with EHRs from 2009 to 2015 (\cite{stats}). \textcolor{red}{maybe put HealthIT US map that shows hospitals who received incentive payment?} While EHRs are widely implemented in hospitals by 2015, physicians still continue to report frustration. "Some healthcare professionals continue to develop workarounds and rely on paper alternatives rather than using EHRs" (\cite{evans2016electronic}).

@@

Using various linked datasets, I construct a physician-level panel spanning from 2009-2017 which measures physicians' exposure to EHRs over time and other relevant characteristics. The different data sets used to construct this panel are described below.

\subsection{MD-PPAS}

Medicare Data on Provider Practice and Specialty is a database of physicians designed to record information about claim counts and specialty from 2009-2017. The relevant information included is physician NPI, primary and secondary specialties, patient counts in up to 12 different zip codes, and fraction of patients in different settings such as office, inpatient or emergency room. First, I make limitations to the sample of physicians based on specialty type. Primary care physicians are the most likely to be exposed to the administrative burden of EHRs in hospitals. Thus, I only include physicians whose primary specialty is hospitalist, internal medicine, pediatrics, general practice or family practice in at least one year (excluding physicians who have specialist listed for all other years). Since the focus of the analysis is physicians working in hospitals, I also drop physicians with less than 20\% of patients in a hospital setting in all years. 
The dependent variables are constructed using MD-PPAS claim and patient count information. Retirement is an indicator variable that is equal to 1 in the first year that all future claims in all zip codes are equal to 0. I do not observe retirement in 2017, and drop those who "retire" in 2016 since one year of zero claims does not necessarily mean retirement.

\subsection{Shared Patient}

CMS Shared Patient data records the number of patients who bill two entities under Medicaid in the same day. For example, if a Medicaid patient is referred to a specialist by a primary care physician, then those two physicians have a shared patient in common. I limit the entities by NPI tax-code to only include shared patients for physician-hospital pairs. I care specifically about physicians who have a close working relationship with hospitals, preferably those who do rounds within at least one hospital. To achieve this, I limit the qualifications for a sufficient physician-hospital pair. First, I only consider doctors who have a primary care role, excluding any specialists. The reason for this is that a primary care physician who does rounds in a hospital is likely to be primarily employed by that hospital (not brought in from a practice). Another reason for this is that primary care doctors interact directly with EHRs during checkups, unlike a surgeon. Second, I limit pairs to those which have a substantial number of patients billed together in the same day. This is to avoid including the potential physician who works in an office but happens to send a small number of patients to a nearby hospital. I consider a substantial number of patients to be at least 30 patients bill both the physician and the hospital in the same day in at least one year, \textcolor{red}{where the exact number chosen is explored in the appendix}. The physicians remaining in the data are likely doing a number of rounds physically inside hospitals, since it is unlikely that many patients will see a primary care doctor in a private office and then immediately go to a hospital in the same day.

@@

It is not uncommon in developed countries for workers to plan for retirement well in advance. Many workers have a retirement age set and spend their working life saving to get to that point. An exogenous shock typically will not change this retirement age because of the large sum of money necessary to formally leave the labor force. However, physicians typically make this decision differently than the majority of workers. Most physicians plan to retire at age 60, but do not actually retire until age 69 (\cite{collier2017challenges}). The reason being that the type of person who selects into being a physician will, on average, have a high regard for helping people. When it comes time to retire, they hesitate to abandon patients they have seen over the course of their career. This decision to delay retirement is not financial. Thus, retiring may be in a physician's choice set years before the realized decision to leave the labor force.

This implies that an exogenous shock to work setting could, if it outweighs the concern for patients, cause an earlier exit. Medical articles suggest that the implementation of EHRs did lead to retirement in older physicians. \textcolor{red}{cite articles}.

Figure \ref{fig:retirefirst} shows the aggregated group time treatment effects of being exposed to an EHR in a hospital. The left panel of the figure includes all physicians and shows a clear increase in the probability of retiring after exposure to an EHR. On the right panel, only physicians over the age of 50 are included. These results are puzzling, as they show that young physicians are the ones "retiring", or leaving the labor force. \textcolor{red}{However, this may just reflect that in later years taking a one or two year break looks like retiring whereas in early years it doesn't. What would cause a physician to drop out of the data for one or two years? Add to the appendix the same analysis but dropping physicians who "retired" after 2014. }

\begin{figure}[ht]
\caption{}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR.pdf}
        \end{minipage}
        \hspace{0.2cm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_old.pdf}
        \end{minipage}
        \label{fig:retirefirst}
\end{figure}

I also repeat this analysis with a more strict version of exposure to EHRs. A physician is only considered exposed if they worked in a low-integration status hospital that implemented an EHR. These results are shown in Figure \ref{fig:retiresecond}.

\begin{figure}[ht]
\caption{}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_li.pdf}
        \end{minipage}
        \hspace{0.2cm}
        \begin{minipage}[b]{0.47\linewidth}
            \centering
            \includegraphics[width=\textwidth]{Objects/ggdid_retire_allEHR_old_li.pdf}
        \end{minipage}
        \label{fig:retiresecond}
\end{figure}

@@

\vspace{5mm}
\begin{figure}[ht]
\caption{Frequency of Data Assistant Enumeration by Year}

    \label{fig:dataassistant_histogram}
\end{figure}
